The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-04 rockerverse.Rmd
addition of the 'PrarieLearn' experience with rocker-pl

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
During the last few years, and coinciding with the rise of R, has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents what we may call the _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

@@

>>>>

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
The last few years, coinciding with the rise of R, have also seen the rise of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents what we may call the _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

@@

Another example of a Docker-based teaching and testing platform is [PrairieLearn](https://github.com/PrairieLearn) which is being developed at the University of Illinois at Urbana-Champaign \citep{prairielearn:2018} where it has already been in extensive use across several faculties (along with initial use on some other campuses).
It uses Docker containers a key components, both internally for its operations (programmed mainly in Python as well as in Javascript), as well as for two reference containers providing, respectively, Python and R autograders. 
A key design decision made by PrairieLearn permits _external_ grading container to be supplied and accessed via a well-defined interface of invoking, essentially, a single script `run.sh` expecting JSON based configuration, relying on a well-defined file layout containing support files, exam questions, supplementary data as well as student submissions, and returning per-question evaluations as JSON result files which are evaluated, aggregated and recorded in SQL. 
The [Data Science Programming Methods](https://stat430.com) course \citep{stat430:2019} uses this via the custom [`rocker-pl`](https://github.com/stat430dspm/rocker-pl) container \citep{rocker-pl:2019}.\footnote{The reference R container was unavailable at the time, and also relies on a heavier CentOS-based build so that a lighter and alternative was established.}
The `rocker-pl` container extends `rocker/r-base` with both the [`plr`](https://github.com/stat430dspm/plr) \citep{pkg:plr:2019} custom R package for integration into PrarieLearn testing and question evaluation, along with the actual R packages used in instruction and testing for the course in question.
As automated grading of submitted student answers is close to the well-understood problem of unit testing, the \CRANpkg{tinytest} package \citep{CRAN:tinytest} is used as it offers both its core features for testing as well as clean extensibility which package \CRANpkg{ttdo} \citep{CRAN:ttdo} utilizes to render differences between incorrect answers and reference answers by invoking the \CRANpkg{diffobj} package \citep{CRAN:diffobj} for context-sensitive colorized displays of these differences.
The setup, described in more detail by \citet{paper:r_autograder}, is an excellent illustration of both the versatility and flexibility offered by Docker-based approaches in teaching and testing.

### 2020-03-04 rockerverse.tex
addition of the 'PrarieLearn' experience with rocker-pl

The R community continues to grow. This can be seen in the number of new
packages on CRAN, which is still on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies that are
adopting R, as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. During the
last few years, and coinciding with the rise of R, has been the advent
of \href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distributing and deploying of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides Docker images with R
(see the next section for more details). The considerable uptake and
continued evolution of the Rocker~Project has led to numerous projects
that extend or build upon Rocker images, ranging from
reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.}
research to production deployments. As such, this article presents what
we may call the \emph{Rockerverse} of projects across all development
stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and
environment with other containerisation solutions. Our main
contributions is a coherent picture of the current status of using
containers in, with, and for R.

@@

\href{https://cyverse.org}{CyVerse} is an open source, NSF-funded
cyberinfrastructure platform for the life sciences providing easy access
to computing and storage resources \citep{merchant_iplant_2016}.
CyVerse's a browser-based `data science workbench' is called
\href{https://cyverse.org/discovery-environment}{Discovery Environment}
(DE). DE uses a combination of
\href{https://research.cs.wisc.edu/htcondor/}{HTCondor} and Kubernetes
for orchestrating container-based analysis and integrates with external
HPC, i.e., \href{https://www.xsede.org/}{NSF-XSEDE}, through
\href{https://www.tacc.utexas.edu/tapis}{TAPIS} API's. In addition,
CyVerse provides a multi-petabyte Data Store based on
\href{https://irods.org/}{iRODS} with shared access by users. The DE
runs Docker containers on demand, with users able to integrate bespoke
containers from DockerHub or other registries
\citep{devisetty_bringing_2016}. Rocker images integration in the DE is
designed to provide life science researchers with scalable,
compute-intensive R analysis capabilities needed for large and complex
datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial
data etc.). These capabilities give users flexibility similar to Binder,
but allows containers to run on larger computational resources (RAM,
CPU, Disk), and for longer periods of time (days to weeks). The Rocker
Project's RStudio and Shiny containers are integrated into the DE by
deriving new images from Rocker
images\footnote{See \href{https://github.com/cyverse-vice/}{https://github.com/cyverse-vice/} for \texttt{Dockerfile}s and configuration scripts; images are auto-built on DockerHub at \href{https://hub.docker.com/u/cyversevice}{https://hub.docker.com/u/cyversevice}.}.
These images include a reverse proxy using \texttt{nginx} to handle
communication with CyVerse's authentication systems, which allow users
to invite other registered users to securely access the same instance
\citep{rstudio_proxy_2020}. The images further include specific tools,
such as the CLI utility \texttt{icommands} to access iRODS. Users can
reference the same base image and add packages to further customize
their R environment, or they can use CyVerse's image registry where a
user's published, i.e., publicly available, containers are held for
additional reproducibility. CyVerse user accounts are free (with some
limitations for non-US users) and the
\href{https://learning.cyverse.org/}{CyVerse Learning Center} allows
community members to document their containers to support training and
education.

@@

>>>>

The R community continues to grow. This can be seen in the number of new
packages on CRAN, which is still on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies that are
adopting R, as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. The last few
years, coinciding with the rise of R, have also seen the rise of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distributing and deploying of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides Docker images with R
(see the next section for more details). The considerable uptake and
continued evolution of the Rocker~Project has led to numerous projects
that extend or build upon Rocker images, ranging from
reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.}
research to production deployments. As such, this article presents what
we may call the \emph{Rockerverse} of projects across all development
stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and
environment with other containerisation solutions. Our main
contributions is a coherent picture of the current status of using
containers in, with, and for R.

@@

\href{https://cyverse.org}{CyVerse} is an open source, NSF-funded
cyberinfrastructure platform for the life sciences providing easy access
to computing and storage resources \citep{merchant_iplant_2016}. CyVerse
has a browser-based `data science workbench' called the
\href{https://cyverse.org/discovery-environment}{Discovery Environment}
(DE). The DE uses a combination of
\href{https://research.cs.wisc.edu/htcondor/}{HTCondor} and Kubernetes
for orchestrating container-based analysis and integrates with external
HPC, i.e., \href{https://www.xsede.org/}{NSF-XSEDE}, through
\href{https://www.tacc.utexas.edu/tapis}{TAPIS}(TACC-API's). CyVerse
hosts a multi-petabyte Data Store based on
\href{https://irods.org/}{iRODS} with shared access by its users. The DE
runs Docker containers on demand, with users able to integrate bespoke
containers from DockerHub or other registries
\citep{devisetty_bringing_2016}. Rocker image integration in the DE is
designed to provide researchers with scalable, compute-intensive, R
analysis capabilities for large and complex datasets (e.g.,
genomics/multi-omics, GWAS, phenotypic data, geospatial data, etc.).
These capabilities give users flexibility similar to Binder, but allow
containers to be run on larger computational resources (RAM, CPU, Disk,
GPU), and for longer periods of time (days to weeks). The Rocker
Project's RStudio and Shiny are integrated into the DE by deriving new
images from Rocker
images\footnote{See \href{https://github.com/cyverse-vice/}{https://github.com/cyverse-vice/} for \texttt{Dockerfile}s and configuration scripts; images are auto-built on DockerHub at \href{https://hub.docker.com/u/cyversevice}{https://hub.docker.com/u/cyversevice}.}.
These new images include a reverse proxy using \texttt{nginx} to handle
communication with CyVerse's authentication system
\citep{rstudio_proxy_2020}; CyVerse also allows owners to invite other
registered users to securely access the same instance. The CyVerse
Rocker images further include tools for connecting to its Data Store,
such as the CLI utility \texttt{icommands} for iRODS. CyVerse accounts
are free (with some limitations for non-US users), and the
\href{https://learning.cyverse.org/}{CyVerse Learning Center} provides
community members with information about the platform, incuding training
and education opportunities.

@@

Another example of a Docker-based teaching and testing platform is
\href{https://github.com/PrairieLearn}{PrairieLearn} which is being
developed at the University of Illinois at Urbana-Champaign
\citep{prairielearn:2018} where it has already been in extensive use
across several faculties (along with initial use on some other
campuses). It uses Docker containers a key components, both internally
for its operations (programmed mainly in Python as well as in
Javascript), as well as for two reference containers providing,
respectively, Python and R autograders. A key design decision made by
PrairieLearn permits \emph{external} grading container to be supplied
and accessed via a well-defined interface of invoking, essentially, a
single script \texttt{run.sh} expecting JSON based configuration,
relying on a well-defined file layout containing support files, exam
questions, supplementary data as well as student submissions, and
returning per-question evaluations as JSON result files which are
evaluated, aggregated and recorded in SQL. The
\href{https://stat430.com}{Data Science Programming Methods} course
\citep{stat430:2019} uses this via the custom
\href{https://github.com/stat430dspm/rocker-pl}{\texttt{rocker-pl}}
container
\citep{rocker-pl:2019}.\footnote{The reference R container was unavailable at the time, and also relies on a heavier CentOS-based build so that a lighter and alternative was established.}
The \texttt{rocker-pl} container extends \texttt{rocker/r-base} with
both the \href{https://github.com/stat430dspm/plr}{\texttt{plr}}
\citep{pkg:plr:2019} custom R package for integration into PrarieLearn
testing and question evaluation, along with the actual R packages used
in instruction and testing for the course in question. As automated
grading of submitted student answers is close to the well-understood
problem of unit testing, the \CRANpkg{tinytest} package
\citep{CRAN:tinytest} is used as it offers both its core features for
testing as well as clean extensibility which package \CRANpkg{ttdo}
\citep{CRAN:ttdo} utilizes to render differences between incorrect
answers and reference answers by invoking the \CRANpkg{diffobj} package
\citep{CRAN:diffobj} for context-sensitive colorized displays of these
differences. The setup, described in more detail by
\citet{paper:r_autograder}, is an excellent illustration of both the
versatility and flexibility offered by Docker-based approaches in
teaching and testing.
